A Spectral Analysis of the Domain Decomposed Monte Carlo Method for Linear Systems

### 2015-04-14 doc/ned2015/slattery_wilson_evans.tex
adding improved data plots

To test the relationships developed by the spectral analysis, we form two
simple numerical experiments using the diffusion model problem: one to measure
the length of the random walks as a function of the iteration matrix
eigenvalues, and one to measure the domain leakage fraction as a function of
the iteration matrix eigenvalues and the discretization properties. Before
doing this, we verify our computation of the spectral radius of the iteration
matrix by numerically computing the largest eigenvalue of the diffusion
operator using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ was used and the
absorption cross varied from 0 to 100 while the scattering cross section was
fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured spectral
radius of the iteration matrix and the computed spectral radius for the
preconditioned diffusion operator using Eq~(\ref{eq:iteration_radius}) as
function of the absorption to scattering ratio $(\Sigma_a /
\Sigma_s)$. Excellent agreement was observed between the analytic and
numerical results with all data points computed within the tolerance of the
iterative eigenvalue solver.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{measured_spec_rad.png}
  \end{center}
  \caption{Measured and analytic preconditioned diffusion operator
    spectral radius as a function of the absorption cross section to
    scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
    $h=1.0$ were used. The red data was computed numerically by an
    eigensolver while the black dashed data was generated by
    Eq~(\ref{eq:iteration_radius}).}
  \label{fig:measured_spec_rad}
\end{figure}

@@

We next create an experiment to measure the length of the random walks
generated by the adjoint Neumann-Ulam solver. To do this, we again use a $100
\times 100$ square grid with $h=0.1$ and the absorption cross varied from 0 to
100 while the scattering cross section was fixed at unity. Three weight cutoff
values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with 10,000 random
walks generated by a point source of strength 1 in the center of the
domain. For each of the random walks, the number of transitions made was
tallied to provide an effective value of $k$ for each random walk. This value
was then averaged over all random walks to get a measured value of $k$ for the
particular operator. On the left, Figure~\ref{fig:measured_length} presents
these measurements as well as the analytic result computed by
Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix spectral
radius, $\rho(\ve{H})$. On the right, Figure~\ref{fig:measured_length} gives
the relative error between the predicted and observed results. We note good
qualitative agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=\textwidth]{measured_length.png}
  \end{center}
  \caption{Measured and analytic random walk length as a function of
    the iteration matrix spectral radius. The weight cutoff was
    varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
    plot, the red data was computed numerically by an adjoint
    Neumann-Ulam implementation while the black dashed data was
    generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
    relative error between the predicted and measured results is
    presented for each weight cutoff.}
  \label{fig:measured_length}
\end{figure}

@@

To compute the leakage fraction numerically, \sn{3}{5} random walks were
sampled from a uniform source of strength unity over the global domain. At the
start of a stage of random walks, the number of random walks starting in the
center domain was computed and as the stage progressed, the number of random
walks that exited that domain was tallied with the ratio of the two numbers
providing a numerical measure of the leakage
fraction. Figure~\ref{fig:measured_leakage} gives the domain leakage
measurements for the domain in the center of the global grid as well as the
analytic result computed by Eqs~(\ref{eq:wigner_domain_leakage}) and
(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration matrix
spectral radius.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.9\textwidth]{leakage_variation.png}
  \end{center}
  \caption{Measured and analytic domain leakage as a function of the iteration
    matrix spectral radius. To test the behavior with respect to domain size,
    $n_i=50$, $n_i=100$,and $n_i=200$ were used. The red data was computed
    numerically by a domain-decomposed adjoint Neumann-Ulam implementation,
    the black dashed data was generated by
    Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
    approximation, and the dashed-dotted black data was generated by
    Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
    approximation.}
  \label{fig:measured_leakage}
\end{figure}
Again, we note good qualitative agreement between the measured and
analytic quantities but we begin to see the limits of the leakage
approximations. To compare the quality of the two approximations, the
absolute error between the computed leakage fraction and that
generated by the Wigner rational and mean chord approximations is
plotted in Figure~\ref{fig:leakage_error} for all domain sizes
tested. 
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.9\textwidth]{leakage_error.png}
  \end{center}
  \caption{Measured and analytic domain leakage absolute error as a
    function of the iteration matrix spectral radius.  To test the
    behavior with respect to domain size, $n_i=50$ (green),
    $n_i=100$ (blue), and $n_i=200$ (red) were used. The dashed
    lines represent the error using the Wigner rational
    approximation while the solid lines represent the error using
    the mean-chord approximation.}
  \label{fig:leakage_error}
\end{figure}
From these error results, the mean chord approximation is shown to have a
lower error for ill-conditioned systems as compared to the Wigner
approximation while the Wigner approximation produces less error for more
well-conditioned systems. We also note that for optically thick domains, the
error is likely corresponded to that observed in
Figure~\ref{fig:measured_length} for the $k$ parameter while the large
relative error in $k$ for optically thin domains does not affect the
approximation significantly. In general, the mean chord approximation is a
better choice to estimate the leakage fraction in a domain from the adjoint
Neumann-Ulam method and except for a single data point with $n_i=50$, the mean
chord approximation yielded leakage fractions within 0.05 of the measured
results. As the domain becomes more optically thick (with both increasing
$n_i$ and decreasing $\rho(\ve{H})$), the approximations are more accurate.

>>>>

To test the relationships developed by the spectral analysis, we form two
simple numerical experiments using the diffusion model problem: one to measure
the length of the random walks as a function of the iteration matrix
eigenvalues, and one to measure the domain leakage fraction as a function of
the iteration matrix eigenvalues and the discretization properties. Before
doing this, we verify our computation of the spectral radius of the iteration
matrix by numerically computing the largest eigenvalue of the diffusion
operator using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ was used and the
absorption cross varied from 0 to 100 while the scattering cross section was
fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured spectral
radius of the iteration matrix and the computed spectral radius for the
preconditioned diffusion operator using Eq~(\ref{eq:iteration_radius}) as
function of the absorption to scattering ratio $(\Sigma_a /
\Sigma_s)$. Excellent agreement was observed between the analytic and
numerical results with all data points computed within the tolerance of the
iterative eigenvalue solver.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{spec_rad_results.eps}
  \end{center}
  \caption{Measured and analytic preconditioned diffusion operator
    spectral radius as a function of the absorption cross section to
    scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
    $h=1.0$ were used. The colored data was computed numerically by an
    eigensolver while the black dashed data was generated by
    Eq~(\ref{eq:iteration_radius}).}
  \label{fig:measured_spec_rad}
\end{figure}

@@

We next create an experiment to measure the length of the random walks
generated by the adjoint Neumann-Ulam solver. To do this, we again use a $100
\times 100$ square grid with $h=0.1$ and the absorption cross varied from 0 to
100 while the scattering cross section was fixed at unity. Three weight cutoff
values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with 10,000 random
walks generated by a point source of strength 1 in the center of the
domain. For each of the random walks, the number of transitions made was
tallied to provide an effective value of $k$ for each random walk. This value
was then averaged over all random walks to get a measured value of $k$ for the
particular operator. Figure~\ref{fig:measured_length} presents
these measurements as well as the analytic result computed by
Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix spectral
radius, $\rho(\ve{H})$. Figure~\ref{fig:measured_length_error} gives
the relative error between the predicted and observed results. We note good
qualitative agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{random_walk_length.eps}
  \end{center}
  \caption{Measured and analytic random walk length as a function of the
    iteration matrix spectral radius. The weight cutoff was varied with
    \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. The colored data was computed
    numerically by an adjoint Neumann-Ulam implementation while the black
    dashed data was generated by Eq~(\ref{eq:analytic_k}).}
  \label{fig:measured_length}
\end{figure}
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{random_walk_error.eps}
  \end{center}
  \caption{Relative error between computed random walk length analytic length
    computed by Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
    spectral radius. The weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4},
    and \sn{1}{-8}.}
  \label{fig:measured_length_error}
\end{figure}

@@

To compute the leakage fraction numerically, \sn{3}{5} random walks were
sampled from a uniform source of strength unity over the global domain. At the
start of a stage of random walks, the number of random walks starting in the
center domain was computed and as the stage progressed, the number of random
walks that exited that domain was tallied with the ratio of the two numbers
providing a numerical measure of the leakage
fraction. Figure~\ref{fig:measured_leakage} gives the domain leakage
measurements for the domain in the center of the global grid as well as the
analytic result computed by Eqs~(\ref{eq:wigner_domain_leakage}) and
(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration matrix
spectral radius.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{leakage_fraction.eps}
  \end{center}
  \caption{Measured and analytic domain leakage as a function of the iteration
    matrix spectral radius. To test the behavior with respect to domain size,
    $n_i=50$, $n_i=100$,and $n_i=200$ were used. The colored data was computed
    numerically by a domain-decomposed adjoint Neumann-Ulam implementation,
    the black dashed data was generated by
    Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
    approximation, and the dashed-dotted black data was generated by
    Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
    approximation.}
  \label{fig:measured_leakage}
\end{figure}
Again, we note good qualitative agreement between the measured and
analytic quantities but we begin to see the limits of the leakage
approximations. To compare the quality of the two approximations, the
absolute error between the computed leakage fraction and that
generated by the Wigner rational and mean chord approximations is
plotted in Figure~\ref{fig:leakage_error} for all domain sizes
tested. 
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{leakage_error.eps}
  \end{center}
  \caption{Measured and analytic domain leakage absolute error as a
    function of the iteration matrix spectral radius.  To test the
    behavior with respect to domain size, $n_i=50$,
    $n_i=100$, and $n_i=200$ were used. The dashed
    lines represent the error using the Wigner rational
    approximation while the solid lines represent the error using
    the mean-chord approximation.}
  \label{fig:leakage_error}
\end{figure}
From these error results, the mean chord approximation is shown to have a
lower error for ill-conditioned systems as compared to the Wigner
approximation while the Wigner approximation produces less error for more
well-conditioned systems. We also note that for optically thick domains, the
error is likely corresponded to that observed in
Figure~\ref{fig:measured_length} for the $k$ parameter while the large
relative error in $k$ for optically thin domains does not affect the
approximation significantly. In general, the mean chord approximation is a
better choice to estimate the leakage fraction in a domain from the adjoint
Neumann-Ulam method and except for a single data point with $n_i=50$, the mean
chord approximation yielded leakage fractions within 0.05 of the measured
results. As the domain becomes more optically thick (with both increasing
$n_i$ and decreasing $\rho(\ve{H})$), the approximations are more accurate.
